---
title: "Credit Balance Prediction Code"
output: html_notebook
---

Setup: Reading in the data, calling packages & setting the seed
We are working with a credit balance data set with 400 observations and 11 variables.
```{r, eval=FALSE}
data <- read.table("Data.txt", header=TRUE)

head(data)

library(MASS)
library(ISLR)
library(leaps)

set.seed(123)
```

Output variable
```{r, eval=FALSE}
summary(data)
str(data)
hist(data$Balance, breaks = 16, xlab = "Balance", main = "Histogram of Balance" )
```

1. Predictor Variables and Measurement Scales 
Having a look at the plots of all variables to learn about their format
```{r, eval=FALSE}
str(data)

plot(data$Income, data$Balance)
plot(data$Limit, data$Balance)
plot(data$Rating, data$Balance)
plot(data$Cards, data$Balance)
plot(data$Education, data$Balance)
plot(data$Age, data$Balance)
plot(data$Gender, data$Balance)
plot(data$Student, data$Balance)
plot(data$Married, data$Balance)
plot(data$Ethnicity, data$Balance)


```

Cards variable 
```{r fig.height=10, fig.width=10, eval=FALSE}
par(mfrow=c(1,2))
plot(data$Cards, data$Balance, main = "Balance and Cards", ylab="Balance", xlab="Cards")
plot(tapply(data$Balance, INDEX=data$Cards, FUN = mean), ylab = "Balance", xlab="Number of Cards", main="Means of Credit Balance")

CardsNew <- ifelse(data$Cards == 1, 1, 2)
CardsNew <- factor(CardsNew)
data <- cbind(data, CardsNew)

plot(data$CardsNew, data$Balance, main = "Balance and Cards", ylab="Balance", xlab="Cards")
```

Education variable 
```{r, eval=FALSE}
#Education years , grouping first few and last few 
plot(data$Education, data$Balance)
plot(tapply(data$Balance, INDEX=data$Education, FUN = mean))

EducationNew <- ifelse(data$Education < 8, 7, data$Education)
EducationNew <- ifelse(EducationNew > 17,18, EducationNew)
dataNew <- data.frame(data, EducationNew)
dataNew

#look if there is a linear trend between the means
plot(dataNew$EducationNew, dataNew$Balance)
plot(tapply(dataNew$Balance, INDEX=dataNew$EducationNew, FUN = mean), type="l")
```

2. Relationiship between the input and output variables

Ploting the pairs 
```{r, fig.height=10, fig.width=10, eval=FALSE}
par(mfrow=c(1,1))
pairs(data[,1:7], pch=21)

cov(data[,1:7], data[,1:7])
```

Separate the data into test and training set
```{r, eval=FALSE}
size.sample <- 300 #~80% 

index <- sample(1:nrow(data), size.sample)
train <- dataNew[index,]
test <- dataNew[-index,]
```

Collinearity
```{r, eval=FALSE}
#collinearity between income and limit
summary(lm(Balance~Limit, data=dataNew))

#Then have a look at Rating and Limit. Strongly correlating values should be 
summary(lm(Limit~Rating, data=dataNew))

#Age & Married
summary(lm(Age~Married, data=dataNew)) #weak relationship 14.4%

#Age & Education
summary(lm(Age~Education, data=dataNew)) #no relationship

#Age & Student
summary(lm(Age~Student, data=dataNew)) #no or very weak relationship

#Student & Education
summary(lm(Education~Student, data=dataNew)) #weak relationship
```



Forward linear regression
```{r, eval=FALSE}
#intercept 
summary(lm(Balance~1, data=train))

#we start with Income because of the corelation in the plot 
summary(lm(Balance~Income, data=train))

#Add limit  / Rating
summary(lm(Balance~Income + Limit, data=train))
summary(lm(Balance~Income + Rating, data=train))

#Add Cards
summary(lm(Balance~Income + Cards, data=train))
summary(lm(Balance~Income + CardsNew, data=train))
AIC(lm(Balance~Income + Cards, data=train))
AIC(lm(Balance~Income + CardsNew, data=train))

#Add age
summary(lm(Balance~Income + CardsNew + Age, data=train))

#Add Education
summary(lm(Balance~Income + CardsNew + Age + Education, data=train))
summary(lm(Balance~Income + CardsNew + Age + EducationNew, data=train))

summary(lm(Balance~Income + CardsNew + Education, data=train))
summary(lm(Balance~Income + CardsNew + EducationNew, data=train))

#Add Gender
summary(lm(Balance~Income + CardsNew + Age + EducationNew + Gender, data=train))
summary(lm(Balance~Income + CardsNew + EducationNew + Gender, data=train))

#Add Married
summary(lm(Balance~Income + CardsNew + Age + EducationNew + Married, data=train))
summary(lm(Balance~Income + CardsNew + EducationNew + Married, data=train))

#Add Ethnicity
summary(lm(Balance~Income + CardsNew + Age + EducationNew + Ethnicity, data=train))
summary(lm(Balance~Income + CardsNew + EducationNew + Ethnicity, data=train))

#Add Student
summary(lm(Balance~Income + CardsNew + Age + EducationNew + Student, data=train))
summary(lm(Balance~Income + CardsNew + EducationNew + Student, data=train))

summary(lm(Balance~Income + CardsNew + Age + Student, data=train)) #seems to fit the model best 
summary(lm(Balance~Income + CardsNew + Student, data=train))

#Check the model without the CardsNew Variable 
summary(lm(Balance~Income + Age + Student, data=train))
```

3.	Model Setup

model.forward
```{r, eval=FALSE}
X1 <- test$Income
X12 <- test$CardsNew
X5 <- test$Age
X8 <- test$Student

model.forward <- function(X1, X12, X5, X8){
  Y = 4.937693 + 0.152518 * X1 + 0.429402 * X12 +  0.018844 * X5 + 0.991921 * X8
}
```

model.mixedsel
```{r, fig.height=10, fig.width=10, eval=FALSE}
model.mixedsel <- function(X1, X5, X8){
  Y = 5.241914 +  0.152686 * X1 + 0.019869 * X5 + 0.980487 * X8
}

```

model.multiple
```{r, eval=FALSE}
lm.fit <- lm(Balance~.,data=train) #simple linear regression
summary(lm.fit)

summary(lm(Balance~Income + CardsNew + Age + Student + Married, data=train))

model.multiple <- function(X1, X5, X8, X12){
  Y = 4.5380534 +  0.1516959 * X1 + 0.0181150 * X5 + 0.9340799 * X8 + 0.6483078 *X12
}
```


4.	Model accuracy and performance
The two linear regression models we are comparing are: 
```{r, eval=FALSE}
lm.forward <- lm(Balance~Income + CardsNew + Age + Student, data=train)
lm.mixedsel<- lm(Balance~Income + Age + Student, data=train)

```

Plotting the predictions 
```{r fig.height=10, fig.width=10, eval=FALSE}
par(mfrow=c(2,2))
#plot for model.forward
pred.forward <- predict(lm.forward, newdata=test)
plot(pred.forward , test$Balance, ylab = "Balance Values in Testset", xlab = "Predicted Values", main = "Real Values & Predictions (Forward Model)")
abline(c(0,0), 1, col = "red")

#plot for model.mixedsel
pred.mixedsel <- predict(lm.mixedsel, newdata=test)
plot(pred.mixedsel, test$Balance, ylab = "Balance Values in Testset", xlab = "Predicted Values", main = "Real Values & Predictions (Mixedsel Model)")
abline(c(0,0), 1, col = "red")

#residuals plots 
plot(predict(lm.forward), residuals(lm.forward), main="Residuals (Forward Model)", ylab="Residuals", xlab="Predicted Values")
abline(h=0, col = "blue")

plot(predict(lm.mixedsel), residuals(lm.mixedsel), main="Residuals (Mixedsel Model)", ylab="Residuals", xlab="Predicted Values")
abline(h=0, col = "blue")
```

```{r, eval=FALSE}
confint(lm.forward)
confint(lm.mixedsel)

AIC(lm.forward)
AIC(lm.mixedsel)
```

Subset selection for multiple regression 

```{r, eval=FALSE}
regfit.full <- regsubsets(Balance~., train)
reg.summary <- summary(regfit.full)

which.max (reg.summary$adjr2)

par(mfrow =c(2,2))
plot(reg.summary$rss ,xlab=" Number of Variables ",ylab=" RSS",
type="l")

plot(reg.summary$adjr2 ,xlab =" Number of Variables ",
ylab=" Adjusted RSq",type="l")

points (7, reg.summary$adjr2[7], col ="red",cex =2, pch =20)

plot(reg.summary$cp ,xlab =" Number of Variables ",ylab="Cp",
type="l")
which.min (reg.summary$cp )
points (5, reg.summary$cp [5], col ="red",cex =2, pch =20)
which.min (reg.summary$bic )

plot(reg.summary$bic ,xlab=" Number of Variables ",ylab=" BIC",
type="l")
points (3, reg.summary$bic [3], col =" red",cex =2, pch =20)
```


## References
Introduction to Statistical Learning by Gareth James, Daniela Witte, Trevor Hastie and Robert Tibshirani






























